Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-10-12 paper/manuscript.Rmd
work on edits in response to reviewers

---
title: "Eleven years of student replication projects provide evidence on the correlates of replicability in psychology"
author:
  - name: Veronica Boyce
    affiliation: Stanford
    footnote:
      - corresp
  - name: Maya Mathur
    affiliation: Stanford
  - name: Michael C. Frank
    affiliation: Stanford
address:
  - code: Stanford
    address: Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251MA.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: dmk-format.csl # Use any CSL style. See https://www.zotero.org/styles for a good list. Ignored if citation_package: natbib
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
#   - tablesfirst # Default
#   - nomarkers # Default
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt

@@

knitr::opts_chunk$set(echo=F, warning=F, message=F)#dev = "png", dev.args = list(type = "cairo-png")
 library("papaja")
library("bookdown")
library("rticles")
 library(here)
 library(tidyverse)
 library(brms)
library(tidybayes)
library(kableExtra)
library(viridis)
library(cowplot)
library(ggthemes)
#r_refs("r-references.bib", append=F)
theme_set(theme_bw())
options(knitr.table.format = "pdf")

@@

# sample size considerations

# ggplot(d, aes(x=target_N, y=replication_N, color=as.factor(sub_rep)))+geom_point()+coord_trans(x="log", y="log")+geom_abline(slope=1)+ scale_color_viridis(discrete=T, direction = 1, 
#                       name = "Subjective\nReplication\nSuccess\n")
# d |> mutate(bigger_orig=replication_N<.9*target_N,
#             bigger_rep=replication_N>1.1*target_N,
#             similar=!bigger_orig & !bigger_rep) |> 
#   group_by(bigger_orig, bigger_rep, similar) |> tally()

@@

Across the 176 studies, the average subjective replication score was `r round(pluck(sub_rep_rate,1))`%, which we can interpret as an overall subjective replication rate. `r round(pluck(pred_int_rate,1))`% (`r predIntcount`/`r predInttotal`) of replications has outcomes with point-estimates within the prediction interval of the original outcome. The median *p*-original value on the original and replication point-estimates coming from the same distribution was `r round(pluck(porig_rate, 1),2)`, representing the median probability that a replication study's estimate would be at least as extreme as was actually observed, if in fact the replication and original were statistically consistent.

Figure \ref{fig:smd} shows the relationship between original standardized effect size, replication effect size, and subjective replication score. Some studies replicated with similar size effects to the original, and others failed to replicate, with replication effect sizes near zero. On average, there was a diminution of effect sizes from original to replication. This pattern of results is consistent with the results of RP:P [@openscienceconsortium2015].

@@

## Bivariate correlates of replication success

@@

Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented in the replication sample, and the cognitive psychology studies replicated more often than social psychology studies (mean subjective replication scores: `r pluck(cog,2,1) |> round(2)` versus `r pluck(cog,2,4) |> round(2)`). Within and between participants designs were both common, and within-participants designs replicated more often than between-participants designs (mean scores: `r pluck(within,2,2) |> round(2)` versus `r pluck(within,2,1) |> round(2)`). Studies with multiple vignettes replicated more often than single vignetted studies (mean scores: `r pluck(vignette,2,1) |> round(2)` versus `r pluck(vignette,2,2) |> round(2)`). However, there were strong correlations among these experimental features as well as between these experimental features and specific subfields.

Studies with open data, which almost always also had open materials as well, tended to replicate more than studies without open data. Nearly all replication studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person).  While online studies in general show comparable results to studies conducted in person [@crump2013], switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well-adapted (e.g., inductions might have been weaker or instructions might have been insufficient). These factors -- open materials, open data, and online samples -- are more common in more recent studies, and so these effects may partially reflect temporal trends.

@@

# Methods
Our pre-registration, code, and coded data are available at https://osf.io/xwn9m/. <!--TODO VIEW ONLY https://osf.io/xwn9m/?view_only=ebb6df1ae97045cfa5154d6119ec20b1-->

@@

The dataset of replication projects comes from class projects conducted in PSYCH 251 (earlier called PSYCH 254) a graduate-level experimental methods class taught at Stanford by MCF from 2011 to 2022. This class is commonly taken by first year graduate students in psychology and related disciplines, and it has been a requirement of the Psychology PhD since around 2015. Each student chose a study to replicate, implemented the study, wrote analysis code, pre-registered their replication, ran the study, and turned in a structured final report including methods, analytic plan, changes from the original study, confirmatory and exploratory analyses, and discussion of outcomes. Students were encouraged to do experimental replications, but some students chose to replicate correlational outcomes or do computational reproducibility projects instead.  We cannot include the full student reports for confidentiality reasons, but we include over 50 reports that we received permission to share and the template given to students at https://osf.io/xwn9m/. <!-- TODO VIEW ONLY-->

@@

Four of the replication projects were included in RP:P, and 10 were previously reported in @hawkins2018 (which reported 11 student replications from the 2015-2016 class, one of those was excluded from the current sample for being non-experimental).

@@

The subjective replication scores were coded on [0, .25, .5, .75, 1], and we ramapped these to 1-5 to run an ordinal regression predicting replication score. We ran logistic regressions predicting prediction interval and linear regressions predicting $p$-original.

>>>>

---
title: "Eleven years of student replication projects provide evidence on the correlates of replicability in psychology"
author:
  - name: Veronica Boyce
    affiliation: Stanford
    footnote:
      - corresp
  - name: Maya Mathur
    affiliation: Stanford
  - name: Michael C. Frank
    affiliation: Stanford
address:
  - code: Stanford
    address: Stanford University
footnote:
  - code: corresp
    text: "Corresponding author. Email: vboyce@stanford.edu"
bibliography: ["251MA.bib"] # Replace with one or more of your own bibtex files. Better BibTeX for Zotero is your friend
csl: vancouver.csl # Use any CSL style. See https://www.zotero.org/styles for a good list. Ignored if citation_package: natbib
link-citations: TRUE
output:
  bookdown::pdf_document2:
    toc: FALSE
    keep_tex: TRUE
    template: generic_article_template.tex
    #md_extensions: "-autolink_bare_uris"
    number_sections: TRUE
    citation_package: default # Can also be "natbib"
lang: en # Main document language in BCP47 format
geometry: "margin=25mm"
papersize: a4
#linestretch: 2 # for double spacing
endfloat: FALSE # Set to TRUE to turn on latex endfloat package to place figures and tables at end of document
# endfloatoption: # See endfloat documentation for more possibilities
#   - tablesfirst # Default
#   - nomarkers # Default
numberlines: FALSE
authblk: TRUE # FALSE = author affiliations in footnotes; TRUE = author affiliations in a block below author names
footnotehyper: FALSE # TRUE will give you enhanced table footnote capabilities. Set to FALSE to be able to use French blocks. Needed due to what appears to be a latex bug.
urlcolor: blue
linkcolor: blue
citecolor: blue
graphics: TRUE # Needed to be able to include images
tables: TRUE # Needed to be able to include tables
# fancyhdr:
#   first:
#     #headleft: "REPORT-NO-XXXX"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 0pt
#     #footleft: A left foot
#     footrulewidth: 0pt
#   subsequent:
#     #headleft: "NEXT-PAGE-HEADER-LEFT"
#     headright: "Kaplan et al. (2021)"
#     headrulewidth: 1pt
#     footrulewidth: 0pt

@@

knitr::opts_chunk$set(echo=F, warning=F, message=F)#dev = "png", dev.args = list(type = "cairo-png")
 library("papaja")
library("bookdown")
library("rticles")
 library(here)
 library(tidyverse)
 library(brms)
library(tidybayes)
library(kableExtra)
library(viridis)
library(cowplot)
library(ggthemes)
library(reshape2)
#r_refs("r-references.bib", append=F)
theme_set(theme_bw())
options(knitr.table.format = "pdf")

@@

# correlations between outcome measures

for_cor <- d %>% filter(include %in% c("stats", "pred_int")) %>% 
  mutate(predInt=as.numeric(predInt))
sub_pred <- cor.test(for_cor$sub_rep, for_cor$predInt)$estimate
sub_p <- cor.test(for_cor$sub_rep, for_cor$p_orig)$estimate
pred_p <- cor.test(for_cor$predInt, for_cor$p_orig)$estimate

@@

Across the 176 studies, the average subjective replication score was `r round(pluck(sub_rep_rate,1))`%, which we can interpret as an overall subjective replication rate. `r round(pluck(pred_int_rate,1))`% (`r predIntcount`/`r predInttotal`) of replications had outcomes with point-estimates within the prediction interval of the original outcome. The median *p*-original value on the original and replication point-estimates coming from the same distribution was `r round(pluck(porig_rate, 1),2)`, representing the median probability that a replication study's estimate would be at least as extreme as was actually observed, if in fact the replication and original were statistically consistent. The subjective replication scores were moderately correlated with both prediction interval (r=`r round(sub_pred,2)`) and *p*-original (r=`r round(sub_p, 2)`) values; prediction interval and *p*-original were highly correlated with each other (r=`r round(pred_p,2)`).

```{r}
medians <- d %>% filter(include =="stats") %>%
  mutate(difference=target_d_calc-rep_d_calc) %>% summarize(orig=quantile(target_d_calc), rep=quantile(rep_d_calc), diff=quantile(difference))

original_50  <- pluck(medians, 1,3) %>% round(2)
original_25 <- str_c(pluck(medians, 1,2) %>% round(2)," - ",pluck(medians, 1,4) %>% round(2))

rep_50  <- pluck(medians, 2,3) %>% round(2)
rep_25 <- str_c(pluck(medians, 2,2) %>% round(2)," - ",pluck(medians, 2,4) %>% round(2))

diff_50 <- pluck(medians, 3,3) %>% round(2)
diff_25 <- str_c(pluck(medians, 3,2) %>% round(2)," - ",pluck(medians, 3,4) %>% round(2))

medians_successful <- d %>% filter(include =="stats") %>%
  filter(sub_rep %in% c(1)) %>% 
    mutate(diff=target_d_calc-rep_d_calc) %>%
  summarize(orig=quantile(target_d_calc), rep=quantile(rep_d_calc),
            diff=quantile(diff))

original_50_succ  <- pluck(medians_successful, 1,3) %>% round(2)
original_25_succ <- str_c(pluck(medians_successful, 1,2) %>% round(2)," - ",pluck(medians_successful, 1,4) %>% round(2))

rep_50_succ <- pluck(medians_successful, 2,3) %>% round(2)
rep_25_succ <- str_c(pluck(medians_successful, 2,2) %>% round(2)," - ", pluck(medians_successful, 2,4) %>% round(2))

diff_50_succ <- pluck(medians_successful, 3,3) %>% round(2)
diff_25_succ <- str_c(pluck(medians_successful, 3,2) %>% round(2)," - ",pluck(medians_successful, 3,4) %>% round(2))

n_successful <- d %>% filter(include =="stats") %>%
  filter(sub_rep %in% c(1)) %>% nrow()
```

Figure \ref{fig:smd} shows the relationship between original standardized effect size, replication effect size, and subjective replication score. Some studies replicated with similar size effects to the original, and others failed to replicate, with replication effect sizes near zero.

On average, there was a diminution of effect sizes from original to replication. This pattern of results is consistent with the results of RP:P [@openscienceconsortium2015]. Specifically, the median original effect size (in SMD units) was `r original_50` (interquartile range: `r original_25`), and the median replication effect size was `r rep_50` (`r rep_25`), for a median difference of `r diff_50` (`r diff_25`). Among the pairs with a subjective replication score of 1 (N=`r n_successful` with standardized effect sizes), the median original effect size was `r original_50_succ` (interquartile range: `r original_25_succ`), and the median replication effect size was `r rep_50_succ` (`r rep_25_succ`), for a median difference of `r diff_50_succ` (`r diff_25_succ`). Thus, for the highest subjectively-rated replications, effect sizes for original and replication were similar in magnitude on average, with some replications having larger effects than their originals.

@@

```{r}
library(gtsummary)
theme_gtsummary_language("en", big.mark = "")
library(kableExtra)

desc <- d %>% mutate(target_d_calc=ifelse(include=="stats", target_d_calc, NA),
             target_p_calc=ifelse(include=="stats",target_p_calc, NA),
      switch_modality=ifelse(target_on_turk!=replication_on_turk,1,0),
         ss_ratio=replication_N/target_N)%>% 
  select("subfield","open_data", "open_materials","switch_modality",  "stanford_internal","within_between","single_vignette","repeated_measure","target_year","target_N","replication_N", "ss_ratio", "target_d_calc", "target_p_calc") %>% 
  mutate(subfield=factor(subfield, levels=c("cognitive", "social", "other-psych", "non-psych"), labels=c("Cognitive", "Social", "Other psych", "Non-psych")),
         within_between=ifelse(within_between=="between", 0, 1)) %>% 
             tbl_summary(
               label=list(
                 subfield ~ "Subfield",
                 stanford_internal ~ "Original authors at Stanford",
                 target_year~ "Publication Year",
                 single_vignette ~ "Single vignette",
                 open_data ~ "Open data",
                 open_materials ~ "Open materials",
                 within_between ~ "Within subjects design",
                 repeated_measure ~ "Number of trials",
                 target_N ~ "Original sample size",
                 replication_N ~ "Replication sample size",
                 ss_ratio ~ "Ratio of replication/original sample sizes",
                 target_d_calc ~"Original effect size (SMD)",
                 target_p_calc ~ "Original p-value",
                 switch_modality ~ "Switch from in-person to online"
               ),
               statistic = list(all_continuous() ~ "{mean} ({sd})", all_continuous() ~ "{median} ({p25}, {p75})"),
               digits=list(target_year ~ 0, target_p_calc ~4),
    missing="no",
             ) %>% modify_header(label="Feature", stat_0="Summary") %>% 
    modify_footnote(update = everything() ~ NA)

as_kable_extra(desc,format="latex", position="!h", caption="Descriptive properties of the replication dataset. For categorical features, counts and percentages are given. For continuous features, the median value and interquartile range are given. N=176 except for original effect size and p-value where N=112. ", booktabs=TRUE, linesep="") |> kable_styling(full_width=F, font_size=10)

```
## Bivariate correlates of replication success

@@

```{r corr-plot, out.height="40%", fig.width=8, fig.height=4.5, fig.pos="ht", fig.cap="Correlations between predictors. See TODO Table 1 or Methods for descriptions. "}
dat <-  for_cor %>% mutate(cognitive=ifelse(subfield=="cognitive",1,0)) %>% select(-sub_rep, -subfield)

corr_mat <- round(cor(dat),2)

# Install and load reshape2 package

# reduce the size of correlation matrix
melted_corr_mat <-melt(corr_mat) %>% 
  mutate(value=ifelse(Var1==Var2, NA, value),
         value=ifelse(Var1 %in% c("cognitive", "non_psych", "other_psych", "social") &
                        Var2 %in% c("cognitive", "non_psych", "other_psych", "social"),
                      NA, value)) %>% 
  mutate(Var1=factor(Var1, levels=c("social", "cognitive", "other_psych", "non_psych", "is_within", "single_vignette", "stanford", "change_platform", "open_data", "open_mat", "pub_year", "log_trials", "log_sample", "log_ratio_ss"), labels=c("Social psych", "Cognitive psych", "Other psych", "Non-psych", "Within participants design", "Single vignette ","Stanford affiliated", "Switch to online", "Open data", "Open materials", "Publication year", "Log # of trials", "Log original sample size", "Log ratio rep/original sample sizes")),
         Var2=factor(Var2, levels=c("social", "cognitive", "other_psych", "non_psych", "is_within", "single_vignette", "stanford", "change_platform", "open_data", "open_mat", "pub_year", "log_trials", "log_sample", "log_ratio_ss"), labels=c("Social psych", "Cognitive psych", "Other psych", "Non-psych", "Within participants design", "Single vignette ","Stanford affiliated", "Switch to online", "Open data", "Open materials", "Publication year", "Log # of trials", "Log original sample size", "Log ratio rep/original sample sizes")))

# plotting the correlation heatmap
ggplot(data = melted_corr_mat, aes(x=Var1, y=Var2,
                                   fill=value)) + #scale_fill_viridis()+
  theme_classic()+
  scale_fill_gradient2(low="#763386",high="#FDE725FF", na.value="grey")+
  labs(x="", y="")+
  geom_tile(alpha=.7)+
  theme(axis.text.x = element_text(angle = 40, vjust = 1, hjust=1), legend.position="none")+
  geom_text(aes(label=value), size=3)

```
Distributions of study outcomes across some of these properties are shown in Figure \ref{fig:predictors-graph}. Both social and cognitive psychology studies were well represented in the replication sample, and the cognitive psychology studies replicated more often than social psychology studies (mean subjective replication scores: `r pluck(cog,2,1) |> round(2)` versus `r pluck(cog,2,4) |> round(2)`). Within and between participants designs were both common, and within-participants designs replicated more often than between-participants designs (mean scores: `r pluck(within,2,2) |> round(2)` versus `r pluck(within,2,1) |> round(2)`). Studies with multiple vignettes replicated more often than single vignetted studies (mean scores: `r pluck(vignette,2,1) |> round(2)` versus `r pluck(vignette,2,2) |> round(2)`). However, there were strong correlations among these experimental features as well as between these experimental features and specific subfields (Figure \ref{fig:corr-plot}).

Studies with open data, which almost always also had open materials as well, tended to replicate more than studies without open data. Nearly all replication studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person).  While online studies in general show comparable results to studies conducted in person [@crump2013], switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well-adapted (e.g., inductions might have been weaker or instructions might have been insufficient). These factors -- open materials, open data, and online samples -- are more common in more recent studies, and so these effects may partially reflect temporal trends (see Figure \ref{fig:corr-plot} for the correlations between these factors and publication year).

@@

# Methods
Our pre-registration, code, and coded data are available at https://osf.io/xwn9m/ [@osfdata]. <!--TODO VIEW ONLY https://osf.io/xwn9m/?view_only=ebb6df1ae97045cfa5154d6119ec20b1-->

@@

The dataset of replication projects comes from class projects conducted in PSYCH 251 (earlier called PSYCH 254) a graduate-level experimental methods class taught at Stanford by MCF from 2011 to 2022. This class is commonly taken by first year graduate students in psychology and related disciplines, and it has been a requirement of the Psychology PhD since around 2015. Each student chose a study to replicate, implemented the study, wrote analysis code, pre-registered their replication, ran the study, and turned in a structured final report including methods, analytic plan, changes from the original study, confirmatory and exploratory analyses, and discussion of outcomes. Students were encouraged to do experimental replications, but some students chose to replicate correlational outcomes or do computational reproducibility projects instead.  We cannot include the full student reports for confidentiality reasons, but we include over 50 reports that we received permission to share and the template given to students at https://osf.io/xwn9m/ [@osfdata]. <!-- TODO VIEW ONLY-->

@@

Four of the replication projects were included in RP:P, and 10 were previously reported in @hawkins2018 (which reported 11 student replications from the 2015-2016 class; one of those was excluded from the current sample for being non-experimental).

@@

The subjective replication scores were coded on [0, .25, .5, .75, 1], and we remapped these to 1-5 to run an ordinal regression predicting replication score. We ran logistic regressions predicting prediction interval and linear regressions predicting $p$-original.
